---
title: "Class 10: Halloween Mini-Project"
author: "Tahmid Ahmed"
format: gfm
prefer-html: true
---

Importing the Data

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```
```{r}
nrow(candy)
```

>Q1. How many different candy types are in this dataset?

There are 85 different brands of candies within this dataset. 
>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

There are 38 fruity candy types.

```{r}
candy["Warheads", ]$winpercent
candy["Kit Kat" , ]$winpercent
candy["Tootsie Roll Snack Bars", ]$winpercent
```

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy in the dataset is Warheads, and it's winpercent is 39.01. 
>Q4. What is the winpercent value for “Kit Kat”?

76.77%
>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

49.65%

```{r}
library("skimr")
skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent row has values that go above 1 whereas the other rows range from only 0 to 1.

>Q7. What do you think a zero and one represent for the candy$chocolate column?

I believe a 0 represents a FALSE value where the candy isn't chocalate. A 1 represents TRUE where the candy is chocalate. 

>Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)

ggplot(candy) + aes(winpercent) + geom_histogram(bins = 10)
```

>Q9. Is the distribution of winpercent values symmetrical?

No it is skewed right towards higher winpercents.

>Q10. Is the center of the distribution above or below 50%?

```{r}
median(candy$winpercent)
mean(candy$winpercent)
```


The center of the distribution is above 50%.

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?


```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])
```

The average for chocalate candy is higher.

>Q12. Is this difference statistically significant?

```{r}
t.test(candy$winpercent[as.logical(candy$fruity)], candy$winpercent[as.logical(candy$chocolate)])
```

The difference between the averages of chocalate and fruity candy is statistically significant since the p-value obtained is 2.871e-08 which is much smaller than 0.05.

>Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)
candy %>% arrange(winpercent) %>% head(5)
```

The five least liked candies by winpercent are Nik L Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters. 

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% arrange(desc(winpercent)) %>% head(5)
```

The 5 most liked candies by winpercent are Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat, and Snickers.

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q.16 This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)

ggplot(candy) + 
aes(winpercent, reorder(rownames(candy),winpercent)) + geom_col()

```

Color

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

library(ggplot2)
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```

Now, for the first time, using this plot we can answer questions like:
> Q17. What is the worst ranked chocolate candy?

Sixlets
> Q18. What is the best ranked fruity candy?

Starburst

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

Nik L Nip is in the top 5 most expensive and is the least liked at about 22.45%.

>Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolate, pluribus and bar, fruity and bar, etc.
>Q23. Similarly, what two variables are most positively correlated?

Chocalate and winpercent.

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
plot(pca$x[,c(1,2)])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
#ggplotly was not rendering properly in pdf format, only works in html format
#ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, Hard, and Pluribus. These make sense since these groups are the most unique in comparison to the other groups. Fruity, hard, and pluribus don't share as much with the other groups as others do for example chocolate which shares a lot with many groups. 